---
title: "@TITLE@"
subtitle: "@SUBTITLE@"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
    self_contained: true
  pdf_document:
    toc: true
  word_document:
    toc: true
embed-resources: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5.5)
library(qpmR)
.bundle <- readRDS("@BUNDLE@")
round <- .bundle$round
prev  <- .bundle$compare_to
fc    <- round$forecast
sol   <- round$solution
hv    <- intersect(c(if ("pi4" %in% sol$vars) "pi4" else "pi", "i", "y_gap", "q"),
                   sol$vars)
infl  <- hv[1]
at <- function(v, h) {
  d <- fc$paths[fc$paths$variable == v & fc$paths$h == h, ]
  if (nrow(d)) d$mean else NA_real_
}
h_end <- fc$horizon
h_yr  <- min(4L, h_end)
```

## Executive summary

The forecast is conditioned on data through **`r as.character(round$fit$period[round$fit$n_obs])`**
and runs to **`r fc$periods[h_end]`**.

- Inflation (`r infl`) is projected at **`r sprintf("%.1f", at(infl, h_yr))`%**
  four quarters ahead and **`r sprintf("%.1f", at(infl, h_end))`%** at the end of
  the horizon, against a steady state of `r sprintf("%.1f", sol$ss[[infl]])`%.
- The policy rate reaches **`r sprintf("%.1f", at("i", h_yr))`%** after four
  quarters.
- The output gap is **`r sprintf("%+.1f", at("y_gap", h_yr))`** percentage points
  four quarters ahead.

```{r summary-judgment, results = "asis"}
if (!is.null(fc$judgment) && nrow(fc$judgment)) {
  cat(sprintf("- The projection embeds **%d judgment entr%s** (see the ledger below).\n",
              nrow(fc$judgment), if (nrow(fc$judgment) == 1L) "y" else "ies"))
}
if (!is.null(fc$conditions) && sum(fc$conditions$source == "condition")) {
  cn <- fc$conditions[fc$conditions$source == "condition", ]
  cat(sprintf("- It is conditioned on assumed paths for **%s** (%s).\n",
              paste(unique(cn$variable), collapse = ", "),
              if (isTRUE(fc$anticipated)) "announced" else "unanticipated"))
}
```

## The projection

```{r forecast-fan}
plot(fc, vars = hv)
```

```{r forecast-table}
hs <- unique(pmin(c(1, 2, 4, 8, h_end), h_end))
tab <- data.frame(period = fc$periods[hs])
for (v in hv) tab[[v]] <- round(vapply(hs, function(h) at(v, h), numeric(1)), 2)
knitr::kable(tab, row.names = FALSE,
             caption = "Projected paths (posterior mean of the model forecast)")
```

## Where the economy stands

Latent states inferred from the data by the Kalman smoother: the output
gap, equilibrium processes, and the real rate gap that drives demand.

```{r gaps}
latent <- setdiff(sol$vars, round$observables)
gv <- intersect(c("y_gap", "r_gap", "q_gap", "dy_bar", "r_bar", "q_bar"), latent)
if (length(gv)) plot(round$fit, vars = utils::head(gv, 4))
```

```{r decomposition, fig.height = 4.5}
dv <- if ("pi4" %in% sol$vars) "pi4" else "pi"
dec <- qpm_decompose(round$fit, vars = dv)
n <- round$fit$n_obs
plot(dec, var = dv, periods = max(1L, n - 31L):n)
```

## Judgment

```{r judgment, results = "asis"}
if (is.null(fc$judgment) || !nrow(fc$judgment)) {
  cat("This round is pure model output: no judgment was imposed.\n")
} else {
  j <- fc$judgment[, c("id", "author", "variable", "period", "add", "target",
                       "rationale")]
  j$add <- sprintf("%+.2f", j$add)
  j$target <- round(j$target, 2)
  print(knitr::kable(j, row.names = FALSE,
                     caption = "Judgment ledger: every adjustment, its author and its rationale"))
  cat("\n\nImplied shocks supporting the judgment, in standard deviations:\n\n")
  u <- fc$shocks_implied_std
  mx <- apply(abs(u), 2, max)
  mx <- sort(mx[mx > 1e-8], decreasing = TRUE)
  cat(paste(sprintf("- `%s`: %.2f sd%s", names(mx), mx,
                    ifelse(mx > 2, " **(implausibly large)**", "")),
            collapse = "\n"), "\n")
}
```

```{r revision, results = "asis", fig.height = 4.5}
if (!is.null(prev)) {
  cat("## Revision since the previous round\n\n")
  rev <- try(compare_rounds(prev, round, variables = hv), silent = TRUE)
  if (inherits(rev, "qpm_revision")) {
    d <- as.data.frame(rev)
    d <- d[d$variable == infl, ]
    show <- d[, c("period", "old", "new", "total", "parameters",
                  "data_revisions", "new_data", "conditions", "judgment")]
    show[-1] <- lapply(show[-1], round, 2)
    print(knitr::kable(show, row.names = FALSE,
                       caption = sprintf("Revision of %s: %s to %s", infl,
                                         prev$name, round$name)))
    cat("\n")
    plot(rev, variable = infl)
  } else {
    cat("_The rounds could not be compared:_ ", as.character(rev), "\n")
  }
}
```

## Model and calibration

```{r model-info, results = "asis"}
cat(sprintf("Model: **%s** — %d equations, %d shocks.\n\n",
            round$model$name, length(round$model$equations),
            length(round$model$shocks)))
if (length(round$model$meta$blocks)) {
  cat("Country adaptations applied:\n\n")
  for (b in round$model$meta$blocks)
    cat(sprintf("- **%s**%s\n", b$name,
                if (nzchar(b$description)) paste0(" — ", b$description) else ""))
  cat("\n")
}
p <- round$model$params
knitr::kable(data.frame(parameter = names(p), value = unname(p)),
             row.names = FALSE, caption = "Calibration")
```

## Reproducibility

```{r repro, results = "asis"}
cat(sprintf("- Data vintage: %s to %s (%d quarters), observables: %s.\n",
            as.character(round$fit$period[1]),
            as.character(round$fit$period[round$fit$n_obs]),
            round$fit$n_obs, paste(round$observables, collapse = ", ")))
cat(sprintf("- Filter log-likelihood: %.2f%s.\n", round$fit$loglik,
            if (isTRUE(round$fit$diffuse))
              sprintf(" (diffuse initialization, %d unit roots)", round$fit$n_unit) else ""))
lb <- round$fit$diag$ljung_box
if (any(!is.na(lb)))
  cat(sprintf("- Innovation diagnostics: smallest Ljung-Box p-value %.2f (%s).\n",
              min(lb, na.rm = TRUE), names(lb)[which.min(lb)]))
cat(sprintf("- Produced with qpmR %s on %s.\n", round$qpmR_version, round$created))
cat("\nThis round can be re-run and checked against these numbers with `verify_round()`.\n")
```
