---
title: "A forecast in ten minutes: the canonical QPM in qpmR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A forecast in ten minutes: the canonical QPM in qpmR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5)
```

qpmR implements the class of semi-structural quarterly projection models
(QPM) used in central-bank Forecasting and Policy Analysis Systems: an
IS curve, a hybrid Phillips curve, a forward-looking inflation-targeting
rule, and an exchange-rate block, solved under model-consistent
expectations.

## The canonical model

`qpm_template("bkl")` ships the canonical Berg–Karam–Laxton (IMF
WP/06/80–81) small open economy model with an illustrative
emerging-economy calibration — a 5 percent inflation target and a
positive country risk premium.

```{r}
library(qpmR)
m <- qpm_template("bkl")
m
```

The policy rule responds to expected year-on-year inflation four
quarters ahead, `E(pi4[+4])`; the four-quarter identity uses lags back
to `pi[-3]`. qpmR turns those long leads and lags into auxiliary states
automatically when the model is solved.

## Solving and checking

```{r}
sol <- qpm_solve(m)
sol
```

The Blanchard–Kahn line is a real diagnostic: an indeterminate or
explosive model refuses to solve and the error names the usual economic
cause. Violating the Taylor principle, for instance:

```{r, error = TRUE}
qpm_solve(qpm_calibrate(m, c2 = -0.5))
```

## Monetary transmission

```{r}
ir <- irf(sol, shock = "eps_i", horizon = 16)
ir
plot(ir, vars = c("i", "r", "pi", "y_gap", "q", "pi4"))
```

A policy tightening raises the real rate, opens a negative output gap,
appreciates the currency on impact (UIP), and produces the hump-shaped
disinflation with a mild rebound as policy later eases below neutral —
the standard QPM transmission story.

## History and forecast

Until the Kalman filter arrives in 0.2, initial states come from
simulation:

```{r}
histq <- simulate(sol, nsim = 48, seed = 7, burn = 20)
fc <- qpm_forecast(sol, from = histq, horizon = 12)
fc
plot(fc, vars = c("pi", "i", "y_gap", "q"))
```

The fan bands are analytic, from the forecast-error variance recursion
`V_h = P V_{h-1} P' + Q S Q'`.

## Specification checks

```{r}
qpm_lint(m)
```

## Where this is going

0.2 adds the multivariate Kalman filter (gaps, `r*`, equilibrium RER,
shock decompositions); 0.3 adds conditional forecasts, the judgment
ledger, and forecast rounds; 0.4 Bayesian estimation. See the README
roadmap.
