---
title: "A forecast in ten minutes: the canonical QPM in qpmR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A forecast in ten minutes: the canonical QPM in qpmR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5)
```

qpmR implements the class of semi-structural quarterly projection models
(QPM) used in central-bank Forecasting and Policy Analysis Systems: an
IS curve, a hybrid Phillips curve, a forward-looking inflation-targeting
rule, and an exchange-rate block, solved under model-consistent
expectations.

## The canonical model

`qpm_template("bkl")` ships the canonical Berg–Karam–Laxton (IMF
WP/06/80–81) small open economy model with an illustrative
emerging-economy calibration — a 5 percent inflation target and a
positive country risk premium.

```{r}
library(qpmR)
m <- qpm_template("bkl")
m
```

The policy rule responds to expected year-on-year inflation four
quarters ahead, `E(pi4[+4])`; the four-quarter identity uses lags back
to `pi[-3]`. qpmR turns those long leads and lags into auxiliary states
automatically when the model is solved.

## Solving and checking

```{r}
sol <- qpm_solve(m)
sol
```

The Blanchard–Kahn line is a real diagnostic: an indeterminate or
explosive model refuses to solve and the error names the usual economic
cause. Violating the Taylor principle, for instance:

```{r, error = TRUE}
qpm_solve(qpm_calibrate(m, c2 = -0.5))
```

## Monetary transmission

```{r}
ir <- irf(sol, shock = "eps_i", horizon = 16)
ir
plot(ir, vars = c("i", "r", "pi", "y_gap", "q", "pi4"))
```

A policy tightening raises the real rate, opens a negative output gap,
appreciates the currency on impact (UIP), and produces the hump-shaped
disinflation with a mild rebound as policy later eases below neutral —
the standard QPM transmission story.

## History and forecast

Until the Kalman filter arrives in 0.2, initial states come from
simulation:

```{r}
histq <- simulate(sol, nsim = 48, seed = 7, burn = 20)
fc <- qpm_forecast(sol, from = histq, horizon = 12)
fc
plot(fc, vars = c("pi", "i", "y_gap", "q"))
```

The fan bands are analytic, from the forecast-error variance recursion
`V_h = P V_{h-1} P' + Q S Q'`.

## Specification checks

```{r}
qpm_lint(m)
```

## Real data: Czechia

qpmR ships `czechia`, a quarterly dataset from 1996 in the model's units
(see `?czechia`). With `trends = "rw"` the equilibrium real exchange
rate and potential growth become random walks — the model then has unit
roots and `qpm_filter()` switches to diffuse initialization
automatically.

```{r}
mcz <- qpm_calibrate(qpm_template("bkl", trends = "rw"),
                     pi_tar = 2, istar_ss = 2, pistar_ss = 2,
                     prem_ss = 1, a5 = 0.4)
cz <- czechia[czechia$period >= "1999",
              c("period", "pi4", "i", "q", "dy_obs", "istar", "pistar")]
fit <- qpm_filter(mcz, cz)
fit
```

The smoothed latent states reproduce the known history — the pre-GFC
boom, the 2009 and 2013 recessions, the COVID crater, the trend real
appreciation of the koruna, and the post-GFC fall in potential growth:

```{r}
plot(fit, vars = c("y_gap", "dy_bar", "r_bar", "q_gap"))
```

Historical shock decomposition of the 2021-23 inflation wave:

```{r}
dec <- qpm_decompose(fit)
plot(dec, var = "pi4", periods = (fit$n_obs - 33):fit$n_obs)
```

And a forecast from the smoothed end-of-sample state:

```{r}
fc <- qpm_forecast(qpm_solve(mcz), from = fit, horizon = 12)
plot(fc, vars = c("pi4", "i", "y_gap", "q"))
```

Note the honesty of the diagnostics on real data: the Ljung-Box test
flags autocorrelated exchange-rate innovations (the driftless `q_bar`
random walk absorbs the appreciation trend through its shocks), and the
2022-23 inflation surprises show up as large outliers. That is the
model asking for judgment and recalibration — which is what 0.3 is for.

## Policy analysis: conditions, judgment, rounds

A forecast becomes policy analysis when you can impose assumptions on
it. `qpm_condition()` inverts the model for the shocks behind any
assumed path -- with an explicit `anticipated` switch, because an
announced rate path and a sequence of surprises are different economics:

```{r}
base <- qpm_forecast(qpm_solve(mcz), from = fit, horizon = 12)
hold <- qpm_condition(base,
                      i = stats::setNames(rep(3.5, 4), base$periods[1:4]),
                      anticipated = TRUE, instruments = "eps_i")
hold
```

Judgment is a logged, auditable operation rather than a spreadsheet
tweak:

```{r}
judged <- add_judgment(base, pi4 = stats::setNames(0.4, base$periods[3]),
                       author = "prices desk",
                       rationale = "announced energy-tariff increase")
judgment_log(judged)
```

A forecast round archives the whole pipeline -- model, calibration, data
vintage, filtration, conditioned forecast -- as one replayable object,
and `compare_rounds()` answers the question every chief economist asks:
*why did the forecast move?*

```{r}
czA <- cz[cz$period <= "2025-Q4", ]
rA <- qpm_round("2026-Q1 March", mcz, czA, horizon = 12)
rB <- qpm_round("2026-Q3 September", mcz, cz, horizon = 12)
rB <- add_judgment(rB, pi4 = c("2027-Q1" = 0.4), author = "prices desk",
                   rationale = "announced energy-tariff increase")
rev <- compare_rounds(rA, rB, variables = c("pi4", "i", "y_gap"))
print(rev, variables = "pi4", periods = "2027-Q1")
plot(rev, variable = "pi4")
```

The contributions telescope, so they sum to the total revision exactly,
and the decomposition endpoints are verified against the archived
rounds. `save_round()` / `load_round()` / `list_rounds()` manage the
on-disk round store, with human-readable CSV sidecars for auditing
without R.

## Where this is going

0.4 adds Bayesian estimation (priors, posterior sampling on the filter
likelihood, posterior fans); 1.0 the full FPAS workflow with Quarto
report templates. See the README roadmap.
